---
title: "Reproducible Research: Peer Assessment 1"
author: "Thorsten Gomann"
date: "January, 10th 2016"
output: html_document
html_document: PA1.html
---

## Loading and preprocessing the data
First I load the given data.
Instead of unpacking the zip-file to a separate file I unzip and read in one step:

```{r}
df <- read.csv(unz("activity.zip", "activity.csv"))
```

There's no need for preprocessing the data at this step.

## What is mean total number of steps taken per day?
The following histogram shows the number of days grouped by respective number of steps taken per day.

```{r, fig.height=5, fig.width=5}
sumStepsDay <- tapply(df$steps, df$date, sum)
hist(sumStepsDay, xlab="steps", main="Distribution of Total Steps per Day")
```

Remains to calculate the mean and median:

```{r}
mean(sumStepsDay, na.rm=T)
median(sumStepsDay, na.rm=T)
```

So the mean number of steps per day is `r format(mean(sumStepsDay, na.rm=T), nsmall=2)` and the median is `r median(sumStepsDay, na.rm=T)`.

## What is the average daily activity pattern?
To display the average steps per interval I first use the tapply function to aggregate data, throwing the result to a baseplot:
```{r, fig.height=5, fig.width=5}
meanStepsInterval <- tapply(df$steps, df$interval, mean, na.rm=T)
plot(names(meanStepsInterval), meanStepsInterval, type="l", xlab="Interval", 
     ylab="Average Steps", main="Average Daily Activity Pattern")
```

## Imputing missing values
Which values are missing?
```{r}
summary(df)
```

Dates and Intervals are always provided. Missing values are only in steps and always in a full day (date).
So the next step is to calculate the mean for each time interval (omitting NA's) and merge the data with origin data:
```{r}
dfMeanSteps <- aggregate(x=df$steps, by=list(df$interval), mean, na.rm=T)
names(dfMeanSteps) <- c("interval", "cleanedSteps")
dfCleaned <- merge(df, dfMeanSteps)
```

Then I replace the NA's by the <i>cleaned</i> data:
```{r}
dfCleaned$steps[is.na(dfCleaned$steps)] <- dfCleaned$cleanedSteps[is.na(dfCleaned$steps)]
```

Creating the same histogram as above with cleaned data:
```{r, fig.height=5, fig.width=5}
sumCleanedStepsDay <- tapply(dfCleaned$steps, dfCleaned$date, sum)
hist(sumCleanedStepsDay, xlab="steps", main="Distribution of Total Steps per Day (cleaned data)")
```

Also creating the mean and median:
```{r}
mean(sumCleanedStepsDay)
median(sumCleanedStepsDay)
```
So the mean number of steps per day is `r format(mean(sumCleanedStepsDay), nsmall=2)` and the median is `r format(median(sumCleanedStepsDay), nsmall=2)`.

## Are there differences in activity patterns between weekdays and weekends?
To distinguish between weeend and weekday I use the lubridate-package. Considering 1 as Sunday and 7 as Saturday this creates the allocation to the kind of day.
```{r}
library("lubridate")
dfWeekday <- df
dfWeekday$weekend <- ifelse(wday(ymd(df$date)) %in% c(1,7), "weekend", "weekday")
```
By using dplyr-package for aggregating and lattice for plotting I create an aggregated dataset:
```{r message=FALSE}
library("lattice")
library("dplyr")
agg <- dfWeekday %>%
        select(weekend, interval, steps) %>%
        filter(!is.na(steps)) %>%
        group_by(weekend, interval) %>%
        summarize(steps=mean(steps))
xyplot(steps ~ interval | weekend, agg, type="l", layout=c(1,2), ylab="Number of steps", xlab="Interval")
```

So the difference in activity pattern is, that weekdays starts earlier and weekends seem to be with higher activity over the day.
